---
title: "Dating survey analysis"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: true
author: Rohan Selva-Radov
date: "2024-06-07"
---

Data are from the [2023 American Perspectives Survey](https://www.americansurveycenter.org/download/january-2023-american-perspectives-survey/).
Full GitHub repository is [here](https://github.com/rohanselvaradov/APS_jan23).

```{r setup, include=FALSE}
library(haven)
library(tidyverse)

df <- read_dta("survey.dta")

print_labels_for_all_columns <- function(data_frame) {
  for (colname in names(df)) {
    # Check if the column is a labelled vector
    if ("haven_labelled" %in% class(df[[colname]])) {
      # Get the variable label
      var_label <- attr(df[[colname]], "label")
      if (!is.null(var_label)) {
        cat("Variable name:", colname, "\n")
        cat("Variable label:", var_label, "\n")
      } else {
        cat("Variable name:", colname, "has no label.\n")
      }

      # Get value labels
      val_labels <- levels(haven::as_factor(df[[colname]]))
      if (length(val_labels) > 0) {
        cat("Value labels:\n")
        print(val_labels)
      } else {
        cat("No value labels.\n")
      }
    } else {
      # This column is not a labelled vector
      cat("Column", colname, "is not a labelled vector.\n")
    }
    cat("\n") # Add a newline for readability between columns
  }
}

# Call the function
# print_labels_for_all_columns(df)
```
```{r cleaning, include=FALSE}
df <- df %>%
  filter(agecat == 1) %>%
  select(
    agecat, Q1, Q3, contains("Q4"), Q5, Q6, Q9, Q10, Q13, Q14, Q15, Q17,
    Q21d, Q21e, Q21f, Q21g,
    Q24, Q25, gender, ppeducat, ppeduc5, ppethm, ppinc7,
    ppmarit5, xparty4, urb_sub_rur, xurbanicity, pppa1648,
    ppp20071, ppp20072, partyid4, pppa_lgb, income6cat,
    party, marit3, parent, ideocat3
  )


question_labels <- list(
  Q4a = "Enjoy being single more",
  Q4b = "People aren't interested in me",
  Q4c = "Nobody meets my expectations",
  Q4d = "Difficult to meet people",
  Q4e = "Ended a relationship recently",
  Q4f = "More important priorities",
  Q21d = "It's difficult for men and women to be friends",
  Q21e = "Fewer single men today are interested in marriage",
  Q21f = "Fewer single women today are interested in marriage",
  Q21g = "Marriage is out of date"
)


plot_question_by_gender <- function(df, question_column, title = NULL) {
  # Ensure the question column and gender column are treated as factors
  data <- df %>%
    select({{ question_column }}, "gender") %>%
    filter(!is.na(.[[question_column]])) %>%
    as_factor()

  # Use the provided title or default to the column's label attribute
  plot_title <- (if (!is.null(title)) {
    title
  } else {
    attr(data[[question_column]], "label")
  }) %>%
    str_wrap(width = 60)

  # Generate the plot
  data %>%
    ggplot(aes(x = !!sym(question_column), fill = gender)) +
    geom_bar(position = "dodge") +
    scale_fill_manual(values = c("Female" = "red", "Male" = "blue")) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
      axis.title.x = element_blank()
    ) +
    scale_x_discrete(labels = function(x) str_wrap(x, width = 20)) +
    labs(title = plot_title)
}


plot_question_by_gender_stacked <- function(df, question_column, title = NULL) {
  # Ensure the question column and gender column are treated as factors
  data <- df %>%
    select({{ question_column }}, "gender") %>%
    filter(!is.na(.[[question_column]])) %>%
    as_factor() 


  # Use the provided title or default to the column's label attribute
  plot_title <- (if (!is.null(title)) {
    title
  } else {
    attr(data[[question_column]], "label")
  }) %>%
    str_wrap(width = 60)

  # Generate the plot
  data %>%
    ggplot(aes(x = gender, fill = !!sym(question_column))) +
    geom_bar(position = "fill") +
    scale_y_continuous(labels = scales::percent_format()) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
      axis.title.x = element_blank(),
      legend.title = element_blank()
    ) +
    scale_x_discrete(labels = function(x) str_wrap(x, width = 20)) +
    labs(title = plot_title)
}

plot_question_set_by_gender <- function(df, question_prefix, title = NULL) {
  df %>%
    select(contains(question_prefix), gender) %>%
    as_factor() %>%
    pivot_longer(cols = contains(question_prefix), names_to = "Question", values_to = "Response", values_drop_na = TRUE) %>%
    mutate(Question = map_chr(Question, ~ question_labels[[.]])) %>%
    mutate(Question = str_wrap(Question, width = 30)) %>%
    ggplot(aes(x = Response, fill = gender)) +
    geom_bar(position = "dodge", na.rm = TRUE) +
    facet_wrap(~Question, scales = "free_x") +
    scale_fill_manual(values = c("Female" = "red", "Male" = "blue")) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
      strip.text = element_text(size = 11, face = "italic"),
    ) +
    scale_x_discrete(labels = function(x) str_wrap(x, width = 15)) +
    labs(title = str_wrap(title, width = 60))
}

plot_question_set_by_gender_stacked <- function(df, question_prefix, title = NULL) {
  df %>%
    select(contains(question_prefix), gender) %>%
    as_factor() %>%
    pivot_longer(cols = contains(question_prefix), names_to = "Question", values_to = "Response", values_drop_na = TRUE) %>%
    mutate(Question = map_chr(Question, ~ question_labels[[.]])) %>%
    mutate(Question = str_wrap(Question, width = 30)) %>%
    ggplot(aes(x = gender, fill = Response)) +
    geom_bar(position = "fill", na.rm = TRUE) +
    facet_wrap(~Question, scales = "free_x") +
    scale_y_continuous(labels = scales::percent_format()) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
      strip.text = element_text(size = 11, face = "italic"),
    ) +
    scale_x_discrete(labels = function(x) str_wrap(x, width = 15)) +
    labs(title = str_wrap(title, width = 60))
}

```

**For all of the following, we pre-filter to only consider respondents in the 18-29 age group.**

I present all the graphs first stacked by gender to allow comparisons more easily,
and then the identical analyses side-by-side to visualise distributions better. 

## Stacked bar plots
I think the first question is strangely worded -- if I were dating someone
but it's not a "committed" relationship, I probably wouldn't say that I was "Single
or not currently in a relationship",
but I wouldn't fit into the "In a committed relationship, but not living together" 
category either.

```{r plots1a, fig.width=8, fig.height=6, out.width='80%'}
plot_question_by_gender_stacked(df, "Q1")
```

The next question is only asked to respondents who said
they were single or not in a relationship.

```{r plots2a, fig.width=8, fig.height=6, out.width='80%'}
plot_question_by_gender_stacked(df, "Q3")
```

The next question is only asked to respondents who 
said they were single or not in a relationship, and additionally
are not casually dating anybody.

``` {r plots3a, fig.width=8, fig.height=6, out.width='80%'}
plot_question_set_by_gender_stacked(df, "Q4", "Please indicate if each of the following is a reason or not a reason why you are not currently looking for a relationship or dates")
plot_question_by_gender_stacked(df, "Q13", "Even if you are not currently using an online dating site or a dating app, have you ever used an online dating site or a dating app?")
```

The next two questions are only asked to respondents who said
they had at some point used an online dating site or app.

```{r plots4a, fig.width=8, fig.height=6, out.width='80%'}
plot_question_by_gender_stacked(df, "Q14", "Have you ever gone on a date with someone who you met through an online dating website or  dating app?")
plot_question_by_gender_stacked(df, "Q15", "Overall, would you say your OWN personal experiences with online dating sites or dating apps have been…")
plot_question_by_gender_stacked(df, "Q17", "When it comes to dating today, do you think it is...")
plot_question_set_by_gender_stacked(df, "Q21", "Read each statement and please say whether you agree or disagree with each one")
plot_question_by_gender_stacked(df, "Q24")

```

The next question is only asked to respondents who said they were not straight/heterosexual.

```{r plots5a, fig.width=8, fig.height=6, out.width='80%'}
plot_question_by_gender_stacked(df, "Q25", "How important is being [LESBIAN OR GAY/BISEXUAL/QUEER/ASEXUAL OR AROMANTIC] to you?")
```

## Side-by-side bar plots
I think the first question is strangely worded -- if I were dating someone
but it's not a "committed" relationship, I probably wouldn't say that I was "Single
or not currently in a relationship",
but I wouldn't fit into the "In a committed relationship, but not living together" 
category either.

```{r plots1b, fig.width=8, fig.height=6, out.width='80%'}
plot_question_by_gender(df, "Q1")
```

The next question is only asked to respondents who said
they were single or not in a relationship.

```{r plots2b, fig.width=8, fig.height=6, out.width='80%'}
plot_question_by_gender(df, "Q3")
```

The next question is only asked to respondents who 
said they were single or not in a relationship, and additionally
are not casually dating anybody.

``` {r plots3b, fig.width=8, fig.height=6, out.width='80%'}
plot_question_set_by_gender(df, "Q4", "Please indicate if each of the following is a reason or not a reason why you are not currently looking for a relationship or dates")
plot_question_by_gender(df, "Q13", "Even if you are not currently using an online dating site or a dating app, have you ever used an online dating site or a dating app?")
```

The next two questions are only asked to respondents who said
they had at some point used an online dating site or app.

```{r plots4b, fig.width=8, fig.height=6, out.width='80%'}
plot_question_by_gender(df, "Q14", "Have you ever gone on a date with someone who you met through an online dating website or  dating app?")
plot_question_by_gender(df, "Q15", "Overall, would you say your OWN personal experiences with online dating sites or dating apps have been…")
plot_question_by_gender(df, "Q17", "When it comes to dating today, do you think it is...")
plot_question_set_by_gender(df, "Q21", "Read each statement and please say whether you agree or disagree with each one")
plot_question_by_gender(df, "Q24")

```

The next question is only asked to respondents who said they were not straight/heterosexual.

```{r plots5b, fig.width=8, fig.height=6, out.width='80%'}
plot_question_by_gender(df, "Q25", "How important is being [LESBIAN OR GAY/BISEXUAL/QUEER/ASEXUAL OR AROMANTIC] to you?")
```